Evolution of pre-copulatory and post-copulatory strategies of inbreeding avoidance

### 2017-08-19 ms_pre_post.Rmd
Minor edit

In contrast, individuals' phenotypic values for tendency for polyandry ($P_{p}$) cannot map directly onto their genotypic values ($P_{g}$) because $P_{g}$ can evolve to be negative, but females cannot mate with a negative number of additional males [e.g., @Shuker2007; @Evans2013]. Rather, we considered polyandry "threshold trait", whereby continuous genotypic variation translates into expression of discrete phenotypic value(s) at some threshold [@Lynch1998; @Roff1996; @Roff1998; @Duthie]. Accordingly, we allow individuals' phenotypic values for tendency for polyandry to equal genotypic values ($P_{p} = P_{g}$) if $P_{g} \geq 0$, but set $P_{p} = 0$ if $P_{g} < 0$. A negative $P_{g}$ value therefore generates a female that is phenotypically monandrous, while a positive $P_{g}$ value generates a female that can express some degree of polyandry (see details below). Polyandry is therefore influenced by continuous genetic variation but only expressed when $P_{g} > 0$. One important general property of such threshold traits is that deleterious traits are occasionally expressed despite sustained negative selection because recombination among deleterious alleles can cause the underlying genotypic value to exceed the threshold for expression [@Roff1996; @Roff1998].

@@

The probability that an offspring of female $i$ is sired by any one of her mates $j$ is calculated by assigning a perceived fertilisation quality to each $j$ with respect to $i$, $Q^{f}_{ij}$. Perceived fertilisation quality $Q^{f}_{ij}$ is calculated in the same way as perceived mate quality $Q^{m}_{ij}$, such that if female $i$ has a strategy of post-copulatory inbreeding avoidance ($F_{p}<0$), then the perceived quality of male $j$ is $Q^{f}_{ij} = (-F_{p} \times k_{ij} + 1)^{-1}$. If the female has strategy of post-copulatory inbreeding preference ($F_{p}>0$), then the perceived quality of male $j$ is $Q^{f}_{ij} = F_{p} \times k_{ij} + 1$. A relative quality ($q^{f}_{ij}$) is then calculated for each male by dividing his $Q^{f}_{ij}$ by the sum of the $Q^{f}_{ij}$ values across all of a female's mates. These $q^{f}_{ij}$ values, which lie between zero and one, define the probability of paternity. Females produce $n$ offspring, so a female $i$ samples from her mates $n$ times independently and with replacement with a probability of $q^{f}_{ij}$ for each male $j$ to determine the realised distribution of sires. Offspring have equal probability of being female or male. After offspring production, all female and male adults die so that generations are non-overlapping.

@@

For example, @Pomiankowski1987 identified four cost categories relevant to mating frequency and mate choice, including elevated risks of predation or disease transmission, and time or energy expenditure. Polyandrous females might experience increased risk of disease transmission [@Roberts2015], a cost that would more realistically apply to a female's realised number of mates rather than her tendency for polyandry. Polyandrous females might also risk harm caused by sexual conflict over multiple mating [e.g., @Arnqvist2005; @Parker2006]. Inbreeding theory predicts that males should be more tolerant of inbreeding than females, leading to sexual conflict over inbreeding in mating encounters [@Parker1979; @Parker2006; @Kokko2006; @Duthie2015a]. Future models could therefore explicitly consider sexual conflict over both polyandry and pre-copulatory inbreeding, and hence capture internally-consistent mechanistic costs. Further, sexual conflict might also affect post-copulatory inbreeding avoidance. For example, when female guppies (*Poecilia reticulata*) were artificially inseminated with equal quantitites of sperm from full-siblings and unrelated males, more eggs were fertilised by unrelated males because the velocities of full sibling's sperm were reduced by females' ovarian fluids [@Gasparini2011]. In black field crickets (*Teleogryllus commodus*), females attempt to remove the spermatophores of unwanted males after copulation, and are capable of controlling sperm transfer to spermatheca after copulation occurs [@Bussiere2006; @Tuni2013]. Future models could therefore explicitly incorporate such mechanisms in order to better understand effects of post-copulatory sexual conflict on female and male reproductive strategy evolution.

>>>>

In contrast, individuals' phenotypic values for tendency for polyandry ($P_{p}$) cannot map directly onto their genotypic values ($P_{g}$) because $P_{g}$ can evolve to be negative, but females cannot mate with a negative number of additional males [e.g., @Shuker2007; @Evans2013b]. Rather, we considered polyandry "threshold trait", whereby continuous genotypic variation translates into expression of discrete phenotypic value(s) at some threshold [@Lynch1998; @Roff1996; @Roff1998; @Duthie]. Accordingly, we allow individuals' phenotypic values for tendency for polyandry to equal genotypic values ($P_{p} = P_{g}$) if $P_{g} \geq 0$, but set $P_{p} = 0$ if $P_{g} < 0$. A negative $P_{g}$ value therefore generates a female that is phenotypically monandrous, while a positive $P_{g}$ value generates a female that can express some degree of polyandry (see details below). Polyandry is therefore influenced by continuous genetic variation but only expressed when $P_{g} > 0$. One important general property of such threshold traits is that deleterious traits are occasionally expressed despite sustained negative selection because recombination among deleterious alleles can cause the underlying genotypic value to exceed the threshold for expression [@Roff1996; @Roff1998].

@@

The probability that an offspring of female $i$ is sired by any one of her mates $j$ is calculated by assigning a perceived fertilisation quality to each $j$ with respect to $i$, $Q^{f}_{ij}$. Perceived fertilisation quality $Q^{f}_{ij}$ is calculated in the same way as perceived mate quality $Q^{m}_{ij}$, such that if female $i$ has a strategy of post-copulatory inbreeding avoidance ($F_{p}<0$), then the perceived quality of male $j$ is $Q^{f}_{ij} = (-F_{p} \times k_{ij} + 1)^{-1}$. If the female has a strategy of post-copulatory inbreeding preference ($F_{p}>0$), then the perceived quality of male $j$ is $Q^{f}_{ij} = F_{p} \times k_{ij} + 1$. A relative quality ($q^{f}_{ij}$) is then calculated for each male by dividing his $Q^{f}_{ij}$ by the sum of the $Q^{f}_{ij}$ values across all of a female's mates. These $q^{f}_{ij}$ values, which lie between zero and one, define the probability of paternity. Females produce $n$ offspring, so a female $i$ samples from her mates $n$ times independently and with replacement with a probability of $q^{f}_{ij}$ for each male $j$ to determine the realised distribution of sires. Offspring have equal probability of being female or male. After offspring production, all female and male adults die so that generations are non-overlapping.

@@

For example, @Pomiankowski1987 identified four cost categories relevant to mating frequency and mate choice, including elevated risks of predation or disease transmission, and time or energy expenditure. Polyandrous females might experience increased risk of disease transmission [@Roberts2015a], a cost that would more realistically apply to a female's realised number of mates rather than her tendency for polyandry. Polyandrous females might also risk harm caused by sexual conflict over multiple mating [e.g., @Arnqvist2005; @Parker2006]. Inbreeding theory predicts that males should be more tolerant of inbreeding than females, leading to sexual conflict over inbreeding in mating encounters [@Parker1979; @Parker2006; @Kokko2006; @Duthie2015a]. Future models could therefore explicitly consider sexual conflict over both polyandry and pre-copulatory inbreeding, and hence capture internally-consistent mechanistic costs. Further, sexual conflict might also affect post-copulatory inbreeding avoidance. For example, when female guppies (*Poecilia reticulata*) were artificially inseminated with equal quantitites of sperm from full-siblings and unrelated males, more eggs were fertilised by unrelated males because the velocities of full sibling's sperm were reduced by females' ovarian fluids [@Gasparini2011]. In black field crickets (*Teleogryllus commodus*), females attempt to remove the spermatophores of unwanted males after copulation, and are capable of controlling sperm transfer to spermatheca after copulation occurs [@Bussiere2006; @Tuni2013]. Future models could therefore explicitly incorporate such mechanisms in order to better understand effects of post-copulatory sexual conflict on female and male reproductive strategy evolution.

### 2017-08-19 pre_post_biorxiv.tex
Minor edit

In contrast, individuals' phenotypic values for tendency for polyandry ($P_{p}$) cannot map directly onto their genotypic values ($P_{g}$) because $P_{g}$ can evolve to be negative, but females cannot mate with a negative number of additional males \cite[e.g.,][]{Shuker2007, Evans2013}. Rather, we considered polyandry ``threshold trait'', whereby continuous genotypic variation translates into expression of discrete phenotypic value(s) at some threshold \cite[][]{Lynch1998, Roff1996, Roff1998, Duthie}. Accordingly, we allow individuals' phenotypic values for tendency for polyandry to equal genotypic values ($P_{p} = P_{g}$) if $P_{g} \geq 0$, but set $P_{p} = 0$ if $P_{g} < 0$. A negative $P_{g}$ value therefore generates a female that is phenotypically monandrous, while a positive $P_{g}$ value generates a female that can express some degree of polyandry (see details below). Polyandry is therefore influenced by continuous genetic variation but only expressed when $P_{g} > 0$. One important general property of such threshold traits is that deleterious traits are occasionally expressed despite sustained negative selection because recombination among deleterious alleles can cause the underlying genotypic value to exceed the threshold for expression \cite[][]{Roff1996, Roff1998}.

@@

Phenotypic values of the three reproductive strategy traits each incur set costs that combine to independently increase the probability that a focal female will die before mating. Numerous forms and mechanisms of direct costs of reproductive strategies could be hypothesised and modelled; the most appropriate formulation depends on the question (see Discussion). By allowing costs of the three strategies to be directly and independently controlled, our model facilitates direct comparison of evolution of pre-copulatory versus post-copulatory inbreeding avoidance given known cost asymmetries. Qualitatively, such costs on female survival probability are biologically reasonable. For example, polyandrous females that undertake increased mate searching or mating can experience increased predation risk \cite[e.g.,][]{Rowe1988, Ronkainen1994, Koga1998}. Females that express pre-copulatory choice can increase the risk of mortality due to harm caused by sexual conflict over mating, and also risk complete mating failure \cite[which equates to pre-reproductive mortality in semelparous organisms;][]{Rowe1994, Kokko2013}. Finally, females that express post-copulatory choice can pay up-front energetic costs, which might result in trade-offs with survival due to developing physiological or biochemical mechanisms needed to store sperm and successfully bias fertilisation \cite[e.g.,][]{Gasparini2011, Tuni2013, Fitzpatrick2014}.

@@

We predict evolution of polyandry in populations where inbreeding depression is severe and inbreeding avoidance through post-copulatory mechanisms can also evolve, especially if pre-copulatory inbreeding avoidance is costly (Figure 1). Indeed, post-copulatory inbreeding avoidance has been observed under these conditions in experimental systems across diverse taxa \cite[e.g.,][]{Pizzari2004, Firman2008, Bretman2009, Gasparini2011, Tuni2013, Firman2015}.

@@

For example, \cite{Pomiankowski1987} identified four cost categories relevant to mating frequency and mate choice, including elevated risks of predation or disease transmission, and time or energy expenditure. Polyandrous females might experience increased risk of disease transmission \cite[][]{Roberts2015}, a cost that would more realistically apply to a female's realised number of mates rather than her tendency for polyandry. Polyandrous females might also risk harm caused by sexual conflict over multiple mating \cite[e.g.,][]{Arnqvist2005, Parker2006}. Inbreeding theory predicts that males should be more tolerant of inbreeding than females, leading to sexual conflict over inbreeding in mating encounters \cite[][]{Parker1979, Parker2006, Kokko2006, Duthie2015a}. Future models could therefore explicitly consider sexual conflict over both polyandry and pre-copulatory inbreeding, and hence capture internally-consistent mechanistic costs. Further, sexual conflict might also affect post-copulatory inbreeding avoidance. For example, when female guppies (\textit{Poecilia reticulata}) were artificially inseminated with equal quantitites of sperm from full-siblings and unrelated males, more eggs were fertilised by unrelated males because the velocities of full sibling's sperm were reduced by females' ovarian fluids \cite[][]{Gasparini2011}. In black field crickets (\textit{Teleogryllus commodus}), females attempt to remove the spermatophores of unwanted males after copulation, and are capable of controlling sperm transfer to spermatheca after copulation occurs \cite[][]{Bussiere2006, Tuni2013}. Future models could therefore explicitly incorporate such mechanisms in order to better understand effects of post-copulatory sexual conflict on female and male reproductive strategy evolution.

@@

\bibitem[{Dougherty et~al.(2016)}]{Dougherty2016}
Dougherty, L.~R., Simmons, L.~W. \& Shuker, D.~M. 2016. 
\newblock {Postcopulatory sexual selection when a female mates once}.
\newblock \textit{Anim. Behav.} \textbf{116}:13--16.

@@

@@

\bibitem[{Foerster \textit{et~al.}(2003)Foerster, Delhey, Johnsen, Lifjeld, \&
  Kempenaers}]{Foerster2003}
Foerster, K., Delhey K., Johnsen, A., Lifjeld, J.~T. \& Kempenaers B. 2003.
\newblock {Females increase offspring heterozygosity and fitness through
  extra-pair matings}.
\newblock \textit{Nature} \textbf{425}:714--717.

@@

@@

\bibitem[{Morton et~al.(1956)Morton, Crow \& Muller}]{Morton1956}
Morton, N.~E., Crow, J.~F., \& Muller, H.~J. 1956.
\newblock {An estimate of the mutational damage in man from data on
  consanguineous marriages}.
\newblock \textit{Proc. Natl. Acad. Sci. USA} \textbf{42}:855--863.

@@

@@

@@

>>>>

In contrast, individuals' phenotypic values for tendency for polyandry ($P_{p}$) cannot map directly onto their genotypic values ($P_{g}$) because $P_{g}$ can evolve to be negative, but females cannot mate with a negative number of additional males \cite[e.g.,][]{Shuker2007, Evans2013b}. Rather, we considered polyandry ``threshold trait'', whereby continuous genotypic variation translates into expression of discrete phenotypic value(s) at some threshold \cite[][]{Lynch1998, Roff1996, Roff1998, Duthie}. Accordingly, we allow individuals' phenotypic values for tendency for polyandry to equal genotypic values ($P_{p} = P_{g}$) if $P_{g} \geq 0$, but set $P_{p} = 0$ if $P_{g} < 0$. A negative $P_{g}$ value therefore generates a female that is phenotypically monandrous, while a positive $P_{g}$ value generates a female that can express some degree of polyandry (see details below). Polyandry is therefore influenced by continuous genetic variation but only expressed when $P_{g} > 0$. One important general property of such threshold traits is that deleterious traits are occasionally expressed despite sustained negative selection because recombination among deleterious alleles can cause the underlying genotypic value to exceed the threshold for expression \cite[][]{Roff1996, Roff1998}.

@@

Phenotypic values of the three reproductive strategy traits each incur set costs that combine to independently increase the probability that a focal female will die before mating. Numerous forms and mechanisms of direct costs of reproductive strategies could be hypothesised and modelled; the most appropriate formulation depends on the question (see Discussion). By allowing costs of the three strategies to be directly and independently controlled, our model facilitates direct comparison of evolution of pre-copulatory versus post-copulatory inbreeding avoidance given known cost asymmetries. Qualitatively, such costs on female survival probability are biologically reasonable. For example, polyandrous females that undertake increased mate searching or mating can experience increased predation risk \cite[e.g.,][]{Rowe1988, Ronkainen1994, Koga1998}. Females that express pre-copulatory choice can increase the risk of mortality due to harm caused by sexual conflict over mating, and also risk complete mating failure \cite[which equates to pre-reproductive mortality in semelparous organisms;][]{Rowe1994, Kokko2013}. Finally, females that express post-copulatory choice can pay up-front energetic costs, which might result in trade-offs with survival due to developing physiological or biochemical mechanisms needed to store sperm and successfully bias fertilisation \cite[e.g.,][]{Gasparini2011, Tuni2013, Fitzpatrick2014a}.

@@

We predict evolution of polyandry in populations where inbreeding depression is severe and inbreeding avoidance through post-copulatory mechanisms can also evolve, especially if pre-copulatory inbreeding avoidance is costly (Figure 1). Indeed, post-copulatory inbreeding avoidance has been observed under these conditions in experimental systems across diverse taxa \cite[e.g.,][]{Pizzari2004, Firman2008, Firman2015, Bretman2009, Gasparini2011, Tuni2013}.

@@

For example, \cite{Pomiankowski1987} identified four cost categories relevant to mating frequency and mate choice, including elevated risks of predation or disease transmission, and time or energy expenditure. Polyandrous females might experience increased risk of disease transmission \cite[][]{Roberts2015a}, a cost that would more realistically apply to a female's realised number of mates rather than her tendency for polyandry. Polyandrous females might also risk harm caused by sexual conflict over multiple mating \cite[e.g.,][]{Arnqvist2005a, Parker2006}. Inbreeding theory predicts that males should be more tolerant of inbreeding than females, leading to sexual conflict over inbreeding in mating encounters \cite[][]{Parker1979, Parker2006, Kokko2006, Duthie2015a}. Future models could therefore explicitly consider sexual conflict over both polyandry and pre-copulatory inbreeding, and hence capture internally-consistent mechanistic costs. Further, sexual conflict might also affect post-copulatory inbreeding avoidance. For example, when female guppies (\textit{Poecilia reticulata}) were artificially inseminated with equal quantitites of sperm from full-siblings and unrelated males, more eggs were fertilised by unrelated males because the velocities of full sibling's sperm were reduced by females' ovarian fluids \cite[][]{Gasparini2011}. In black field crickets (\textit{Teleogryllus commodus}), females attempt to remove the spermatophores of unwanted males after copulation, and are capable of controlling sperm transfer to spermatheca after copulation occurs \cite[][]{Bussiere2006, Tuni2013}. Future models could therefore explicitly incorporate such mechanisms in order to better understand effects of post-copulatory sexual conflict on female and male reproductive strategy evolution.

@@

\bibitem[{Dougherty \textit{et~al.}(2016)}]{Dougherty2016}
Dougherty, L.~R., Simmons, L.~W. \& Shuker, D.~M. 2016. 
\newblock {Postcopulatory sexual selection when a female mates once}.
\newblock \textit{Anim. Behav.} \textbf{116}:13--16.

@@

\bibitem[{Evan \& Gasparini(2013)}]{Evans2013b}
Evans, J.~P. \& Gasparini, C. 2013.
\newblock {The genetic basis of female multiple mating in a polyandrous 
  livebearing fish}.
\newblock {\textit{Ecol. Evol.}} \textbf{3}:61--66.

\bibitem[{Evans \& Simmons(2008)}]{Evans2008}
Evans, J.~P. \& Simmons, L.~W. 2008.
\newblock {The genetic basis of traits regulating sperm competition and 
  polyandry: Can selection favour the evolution of good- and sexy-sperm?}
\newblock \textit{Genetica} \textbf{134}:5--19.

@@

\bibitem[{Fitzpatrick \textit{et~al.}(2014)}]{Fitzpatrick2014a}
Fitzpatrick, L.~J., Gasparini, C., Fitzpatrick, J.~L. \& Evans, J.~P. 2014.
\newblock {Male-female relatedness and patterns of male reproductive investment 
  in guppies}
\newblock \textit{Biol. Lett.} \textbf{10}:20140166.

\bibitem[{Foerster \textit{et~al.}(2003)Foerster, Delhey, Johnsen, Lifjeld, \&
  Kempenaers}]{Foerster2003}
Foerster, K., Delhey K., Johnsen, A., Lifjeld, J.~T. \& Kempenaers, B. 2003.
\newblock {Females increase offspring heterozygosity and fitness through
  extra-pair matings}.
\newblock \textit{Nature} \textbf{425}:714--717.

@@

\bibitem[{Guillaume \& Perrin(2006)}]{Guillaume2006}
Guillaume, F. \& Perrin, N. 2006.
\newblock {Joint evolution of dispersal and inbreeding load.}
\newblock \textit{Genetics} \textbf{173}:497--509.

@@

\bibitem[{Morton \textit{et~al.}(1956)Morton, Crow \& Muller}]{Morton1956}
Morton, N.~E., Crow, J.~F., \& Muller, H.~J. 1956.
\newblock {An estimate of the mutational damage in man from data on
  consanguineous marriages}.
\newblock \textit{Proc. Natl. Acad. Sci. USA} \textbf{42}:855--863.

@@

\bibitem[{Shuker \textit{et~al.}(2007)}]{Shuker2007}
Shuker, D.~M., Phillimore, A.~J., Burton-Chellew, M.~N., Hodge, S.~E. \& West, S.~A. 2007.
\newblock {The quantitative genetic basis of polyandry in the parasitoid wasp, 
  \textit{Nasonia vitripennis}}.
\newblock \textit{Heredity} \textbf{98}:69--73.

@@

\bibitem[{Wang \textit{et~al.}(1999)}]{Wang1999}
Wang, J., Hill, W.~G., Charlesworth, D. \& Charlesworth, B. 1999.
\newblock {Dynamics of inbreeding depression due to deleterious mutations in 
  small populations: mutation parameters and inbreeding rate}.
\newblock \textit{Genet. Res.} \textbf{74}:165--178.

@@

\bibitem[{White \textit{et~al.}(2014)}]{White2014}
White, J.~W., Rassweiler, A., Samhouri, J.~F., Stier, A.~C. \& White, C. 2014.
\newblock {Ecologists should not use statistical significance tests to interpret
  simulation model results}.
\newblock \textit{Oikos} \textbf{123}:385--388.
